---
title: Introduction to Bayesian inference
subtitle: End of course data analysis project
author: ---
---

# Vaccination against Varicella
The Center for Disease Control (CDC) reports the vaccination coverage of
Varicella among young children. Varicella, commonly known as chicken-
pox, is a highly contagious viral infection caused by the varicella-zoster virus
(VZV). Vaccination against chickenpox has been highly effective in reducing
the incidence and severity of the disease. In the United States, vaccination
against varicella has been part of the routine childhood immunization sched-
ule since the mid-1990s. Since the vaccine’s introduction, there has been a
dramatic decline in the number of chickenpox cases, hospitalizations, and
deaths associated with the disease. The target for vaccination coverage of
varicella (chickenpox) in the United States, as set by the Centers for Dis-
ease Control and Prevention (CDC), is typically around 90% or higher for
children. This high coverage rate is aimed at achieving herd immunity and
preventing outbreaks of chickenpox within communities.


# Project 1: Insurance
The next table summarizes, based on a survey, the number of children in the
birth cohort 2014-2017 that had at least one dose of the Varicella vaccine. It
gives the number of vaccinated children (Vaccinated) amongst the number
of children in the survey (Sample Size). The information is provided for 5
regions of the US, and split according to insurance status (private insurance, uninsured or any Medicaid).

| Geography | Insurance | Vaccinated | Sample Size |
| :--- | :--- | :--- | :--- |
| North Carolina | Any Medicaid | 380 | 419 |
| North Carolina | Private Insurance Only | 632 | 673 |
| North Carolina | Uninsured | 28 | 34 |
| Georgia | Any Medicaid | 363 | 396 |
| Georgia | Private Insurance Only | 527 | 576 |
| Georgia | Uninsured | 36 | 50 |
| Wisconsin | Any Medicaid | 282 | 332 |
| Wisconsin | Private Insurance Only | 514 | 548 |
| Wisconsin | Uninsured | 16 | 34 |
| Florida | Any Medicaid | 446 | 490 |
| Florida | Private Insurance Only | 588 | 628 |
| Florida | Uninsured | 28 | 39 |
| Mississippi | Private Insurance Only | 400 | 441 |
| Mississippi | Uninsured | 27 | 32 |

# Question 1

::: {.callout}
Derive analytically the posterior of the vaccination coverage per ge-
ography and insurance group. Use a conjugate prior that (1) reflects
no knowledge on the vaccination coverage, and (2) reflects that vac-
cination coverage is typically around 90% or higher. Give posterior
summary measures of the vaccination coverage per geography and in-
surance group. Is the choice of the prior impacting your results?
:::

## Theoretical considerations

The outcome *Vaccinated/Not Vaccinated* follows a Bernouilli distribution with parameter $p$:

$V:$ Vaccination status
$V \in \{0,1\}$
$V \sim \mathcal{Bern}(p)$


It is known from theory that the sum of $n$ $i.i.d$ Bernoulli random variables follows a Binomial distribution. This will be used to model the sample outcome: the number of vaccinated people $V_s$ in a random sample of size $n$:

$$V_s = \sum_i^n V_i \sim \mathcal{Binom}(n, \theta)$$

where $\theta$ is the parameter of interest - the vaccine coverage.

In the course, we saw that the Beta distribution is the conjugate prior for binomially distributed data:

| Distribution | Formula |
| :----------- | :------- |
| Prior | $p(\theta) = \mathcal{Beta}(\alpha, \beta)$ |
| Likelihood | $p(y \mid \theta) = {n \choose k} \theta^k (1 - \theta)^{n-k}$ |
| Posterior | $p(\theta \mid y) = \mathcal{Beta}(\alpha + k, \beta + n - k)$ |

The summary measures for the Beta distribution are defined as follows:

| Summary Measure | Formula |
| :--------------- | :------- |
| Mean  | $\frac{\alpha}{\alpha + \beta}$ |
| Median |  See Note |
| Mode  | $\frac{\alpha - 1}{\alpha + \beta - 2}$ for $\alpha, \beta > 1$ |

Note: The median of the Beta distribution does not have a simple closed form expression. It can be approximated numerically or using statistical software.

```{r}
source("./project.R")
```

## Choice of prior distributions

### (1) No prior knowledge
In order to reflect no prior knowledge on the vaccine coverage, the weakly-informative prior Beta(1,1) will be used,  which is equivalent to the uniform distribution over $[0,1]$.

### (2) Vaccine coverage >90%

For modeling prior knowledge that vaccine coverage is about 90%, we chose the Beta(150, 7) distribution. 


### Comparison of priors

```{r}
plot_priors
```

## Results

```{r}
post_summary %>% knitr::kable(
    caption = "Posterior distribution parameters and summary measures per geography for the two different Beta priors",
    col.names = c(
        "Geography", "Insurance",
        "alpha", "beta", "mean", "mode", "var", "HPD LL", "HPD UL",
        "alpha", "beta", "mean", "mode", "var", "HPD LL", "HPD UL"
    ),
    digits = 2) -> testtbl
testtbl %>%
  add_header_above(header = c(
    " "=2, "Beta(1,1) prior"=7, "Beta(150,7) prior"=7
  )) %>%
  add_header_above(header = c(
    " "=2, "Posterior parameters and summary measures"=14
  ))
```

```{r}
grid.arrange(plots)
```

# Question 2

::: {.callout}
nvestigate whether the vaccination coverage is associated with the in-
surance status using a logistic regression model
Yij ∼Binom(πij ,Nij )
with
logit(πij ) = α0 + α1IAnyMedicaid,ij + α2IUninsured,ij
where i is the location, j is the insurance status, πij is the vaccination
coverage and I. are dummy variables. Assume non-informative priors
for the parameters to be estimated. Write and explain the code in
BUGS language
:::

https://github.com/andrewcparnell/jags_examples/blob/master/R%20Code/jags_logistic_regression.R



```bugs
model
{
  # Likelihood
  for (t in 1:T) {
    y[t] ~ dbin(p[t], K)
    logit(p[t]) <- alpha_0 + alpha_1 * x_1[t] + alpha_2 * x_2[t]
  }

  # Priors
  alpha_0 ~ dnorm(0.0,0.01)
  alpha_1 ~ dnorm(0.0,0.01)
  alpha_2 ~ dnorm(0.0,0.01)
}
```

